---
title: "Process LEHD for Commute Patterns"
author: "Chris Day"
date: "2023-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(foreign)
library(sf)
library(jsonlite)
library(mapview)
library(imputeTS)
library(purrr)
```

## Functions
```{r}
intersect_pct <-function(shape1,shape2){
  int <- st_intersection(shape1, shape2) %>% 
    mutate(intersect_area = st_area(.)) %>%   # create new column with shape area
    dplyr::select(SL_COTAZID,GEOID, intersect_area) %>%   # only select columns needed to merge
    st_drop_geometry()  # drop geometry as we don't need it
  
  # Create a fresh area variable for counties
  shape1.1 <- mutate(shape1, shape_area = st_area(shape1))

  # Merge by county name
  shape1.1 <- merge(shape1.1, int, by = "SL_COTAZID", all.x = TRUE)

  # Calculate coverage
  shape1.1 <- shape1.1 %>% 
   mutate(coverage = as.numeric(intersect_area/shape_area))
  
  shape1.1 %>%
    select(SL_COTAZID,GEOID,intersect_area,coverage)
} 
```

## StreetLight Data
```{r}
sl_hbw <- read_csv("data/data2019/StreetLight/SL_HBW.csv")
sltaz <- st_read("data/data2019/StreetLight/StreetLight_TAZ_2021_09_22/StreetLight_TAZ_2021_09_22.shp") %>%
  st_transform(26912)
bg <- st_read(paste0("data/data",year,"/tl_2020_49_bg/tl_2020_49_bg.shp")) %>%
  st_transform(26912) 
```

## City/Township Data
```{r}
#read in township codes and adjust the shortdesc to make merging easier later on
township_codes <- read_csv("data/data2019/citytownship.csv") %>%
  select(1,2) %>%
  mutate(SHORTDESC = case_when(
    CODE3 == "EMT" ~ "Emigration Canyon",
    CODE3 == "CMT" ~ "Copperton",
    CODE3 == "KMT" ~ "Kearns",
    CODE3 == "MMT" ~ "Magna",
    CODE3 == "WHT" ~ "White City",
    TRUE ~ SHORTDESC
  ))
```

```{r}
# a list of all the city/townships in the wfrc areas
wfrc_towns <- c("AFK","ALA","ALP","BDL","BGM","BNT","BRT","CDF","CEN","CHA","CHL","CLF","CLI","CMT","COA","CWH","DAN","DRA","EAG","ELK","EMT","FAR","FCS","FFD","FRR","FTH","GLA","GOS","GRL","HAR","HDT","HDT","HEB","HER","HGH","HNF","HOL","HOO","HVL","IND","INT","KAY","KMS","KMT","LAY","LEH","LIN","MAP","MID","MLC","MMT","MRG","MSL","MUR","MWY","NOG","NSL","OGD","OKL","ORM","PAY","PGR","PLN","PRK","PRY","PVO","PVW","ROY","RVD","RVT","SAN","SAQ","SAR","SFK","SJC","SLC","SLM","SOG","SPV","SSL","SUN","SWE","SYR","TAY","TOO","UIN","VIN","WAT","WBG","WDL","WEB","WHT","WHV","WIL","WJC","WPT","WVC","WXC")
```

## Geographic Data
```{r}
# block level coordinates
blockcity <- read.dbf("data/data2019/blockcentersWBlockGroup.dbf") %>%
  select(GEOID, GEOID10,x,y) %>% st_as_sf(coords = c("x","y"), crs = 4326) %>%
  st_transform(crs = 4326) %>%   
  st_as_sf() %>%
  st_transform(26912) 
# township shapefile
sf_townships <- st_read("data/data2019/Municipalities_Township/Municipalities_Township.shp") %>%
  st_transform(crs = 4326) %>%   
  st_as_sf() %>%
  st_transform(26912)
#get the city at each block gorups
blockpoint_city <- st_join(blockcity, sf_townships, join = st_within)
```

## Streetlight Data Pre-Processing
```{r}
#mapview(sltaz %>% filter(SL_COTAZID == "1002_0"), col.regions="red") + 
#  mapview(bg)
```

```{r}
sltazbg <- intersect_pct(sltaz,bg)
sltaz_nestbg <- sltazbg %>%
  as_tibble() %>%
  select(SL_COTAZID,GEOID) %>%
  nest(GEOID)
sltaz_nestbgp <- sltazbg %>%
  as_tibble() %>%
  select(SL_COTAZID,coverage) %>%
  nest(coverage)

sltazbgj <- sltazbg %>%
  as_tibble() %>%
  select(-geometry,-intersect_area)
```

```{r}
sl_hbw_1 <- sl_hbw %>%
  left_join(sltaz_nestbg, by = c("origin_zone_name" = "SL_COTAZID")) %>%
  unnest() %>%
  rename("origin_zone_bg" = "GEOID") %>%
  left_join(sltazbgj, by = c("origin_zone_name" = "SL_COTAZID","origin_zone_bg" = "GEOID")) %>%
  rename("origin_zone_cov" = "coverage")
sl_hbw_2 <- sl_hbw_1 %>%
  left_join(sltaz_nestbg, by = c("destination_zone_name" = "SL_COTAZID")) %>%
  unnest() %>%
  rename("destination_zone_bg" = "GEOID") %>%
  left_join(sltazbgj, by = c("destination_zone_name" = "SL_COTAZID","destination_zone_bg" = "GEOID")) %>%
  rename("destination_zone_cov" = "coverage")
```

```{r}
sl_hbw_bg <- sl_hbw_2 %>%
  mutate(hbw_traffic_calibrated_trip_volume = o_d_traffic_calibrated_trip_volume * purpose_hbw_percent) %>%
  mutate(hbw_trip = hbw_traffic_calibrated_trip_volume * origin_zone_cov * destination_zone_cov)

sl_hbw_test <- sl_hbw_bg %>%
  group_by(origin_zone_name,destination_zone_name,hbw_traffic_calibrated_trip_volume) %>%
  summarize(test_hbw_vol = sum(hbw_trip))
```

```{r}
sl_hbw_sum <- sl_hbw_bg %>%
  select(origin_zone_bg,destination_zone_bg,hbw_trip) %>%
  group_by(origin_zone_bg,destination_zone_bg) %>%
  summarize(hbw_trips = sum(hbw_trip))
```


Now that we have a dataset of all the streetlight trips from one block group to another, we can summarize it similarly to how we summarized the LODES data, and determine the number of origins and destinations occurring within each city!

## Streetlight Data Summarized at the Blockgroup/City Level
```{r}
blockgroup_city <- blockpoint_city %>%
  as_tibble() %>%
  select(GEOID,SHORTDESC) %>%
  mutate(GEOID = as.character(GEOID)) %>%
  unique() %>%
  left_join(township_codes, by = c("SHORTDESC" = "SHORTDESC")) %>% 
  mutate(CODE3 = ifelse(CODE3 %in% wfrc_towns, CODE3, "XXX"),
         SHORTDESC = ifelse(CODE3 %in% wfrc_towns, SHORTDESC, "Outside WFRC Area"))
```

```{r}
oside_wfrc <- left_join(sl_hbw_sum,blockgroup_city, by = c("origin_zone_bg" = "GEOID")) %>%
  rename("o_geocode" = "origin_zone_bg", "d_geocode" = "destination_zone_bg")
dside_wfrc <- left_join(sl_hbw_sum,blockgroup_city, by = c("destination_zone_bg" = "GEOID")) %>%
  rename("o_geocode" = "origin_zone_bg", "d_geocode" = "destination_zone_bg")

```

```{r}
oside_wide <- oside_wfrc %>%
  select(-SHORTDESC, -o_geocode) %>%
  group_by(d_geocode, CODE3) %>%
  summarize(SumHBW = sum(hbw_trips)) %>% 
  ungroup() %>%
  rename("GEOID" = "d_geocode")
oside_wide_full <- blockseq %>%
  left_join(oside_wide)%>%
  complete(GEOID,CODE3) %>%
  arrange(CODE3) %>%
  pivot_wider(names_from = CODE3, values_from = SumHBW, values_fn = sum, names_glue = "{CODE3}_o") %>%
  na_replace(0)
oside_wide_ttl <- oside_wide_full %>% as.tibble() %>%
  mutate(TTL_o = rowSums(.[2:99])) %>%
  select(-XXX_o)
```

```{r}
dside_wide <- dside_wfrc %>%
  select(-SHORTDESC, -d_geocode) %>%
  group_by(o_geocode, CODE3) %>%
  summarize(SumHBW = sum(hbw_trips)) %>% 
  ungroup() %>%
  rename("GEOID" = "o_geocode")
dside_wide_full <- blockseq %>%
  left_join(dside_wide)%>%
  complete(GEOID,CODE3) %>%
  arrange(CODE3) %>%
  pivot_wider(names_from = CODE3, values_from = SumHBW, values_fn = sum, names_glue = "{CODE3}_d") %>%
  na_replace(0)
dside_wide_ttl <- dside_wide_full %>% as.tibble() %>%
  mutate(TTL_d = rowSums(.[2:99])) %>% 
  select(-XXX_d)
```


```{r}
# merge home and work based data into single dataframe
o_d_ttl <- left_join(oside_wide_ttl, dside_wide_ttl)
```

```{r}
# calculate a fakeid and calculate a column sum
sl_at_blockgroups <- o_d_ttl %>%
  arrange(GEOID) %>%
  mutate(fakeid = row_number() - 1) %>%
  select(198,1:197) %>%
  arrange(fakeid) %>%
bind_rows(summarise(.,
                    across(where(is.numeric), sum),
                    across(where(is.character), ~"colSUM"))) %>%
  slice(n(),1:(n()-1)) %>%
  mutate(fakeid = ifelse(GEOID == "colSUM", 0,fakeid))
```

```{r}
#write output csv to be used in the python script which create the webapp inputs
write_csv(sl_at_blockgroups, paste0("data/data",year,"/sl_at_blockgroup_level",year,".csv"))
```


